---
title: "Attribute Weights and Attention with Varying Sampling Noise"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Attribute Weights and Attention with Varying Sampling Noise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)
library(patchwork)
library(masc)
```


(A) Development of fixations to the higher weighted attribute over time with a fixed
weight difference of 0.5. MASC predicts that higher sampling noise leads to prolonged
sampling of the higher weighted attribute early on before attention switches to the other
attribute. 

(B) shows the predicted difference in attention as a function of weight difference for
varying levels of sampling noise.

```{r}
# Function to simulate the MASC model behavior with various weight differences and noise levels
simulate_weight_attention_relationship <- function(
    n_trials = 100,                      # Number of trials per condition
    weight_diffs = seq(0, 1, by = 0.05), # Weight differences to test
    noise_levels = seq(0.5, 3, by = 0.5),# Noise levels to test
    max_fixations = 100,                 # Maximum number of fixations
    alpha = 10,                          # High search sensitivity (as in MATLAB)
    delta = 0.05                         # Threshold increase (as in MATLAB)
) {
  # Pre-allocate results data frames
  weight_att_results <- data.frame()
  fixation_development <- data.frame()

  # Loop through noise levels
  for (noise in noise_levels) {
    # Loop through weight differences
    for (w_diff in weight_diffs) {
      # Calculate weights - ensuring they're both positive and sum to 1
      # When w_diff = 0, weights are equal (0.5, 0.5)
      # When w_diff = 1, weights are (1, 0) - not allowed, so we'll use (0.99, 0.01)
      w1 <- 0.5 + w_diff/2
      w2 <- 1 - w1

      # Ensure weights are positive and not exactly 0 (use small positive value)
      if (w2 <= 0) {
        w1 <- 0.99
        w2 <- 0.01
      }
      weights <- c(w1, w2)

      # Verify weights sum to 1
      stopifnot(abs(sum(weights) - 1) < 1e-10)

      # Run simulation
      sim <- rMASC(
        n = n_trials,
        n_options = 2,
        n_attributes = 2,
        w = weights,
        sigma = noise,
        alpha = alpha,
        delta = delta,
        max_steps = max_fixations
      )

      # Calculate attention metrics for each trial
      trial_att_diffs <- map_dbl(sim$raw, function(trial) {
        fix_seq <- trial$fix_sequence

        # Count fixations to each attribute
        att_indices <- ceiling(fix_seq / 2)
        att1_fixes <- sum(att_indices == 1)
        att2_fixes <- sum(att_indices == 2)
        total_fixes <- length(fix_seq)

        # Calculate attention difference
        att_diff <- (att1_fixes/total_fixes) - (att2_fixes/total_fixes)
        return(att_diff)
      })

      # Add to weight-attention results
      weight_att_results <- bind_rows(
        weight_att_results,
        tibble(
          noise_level = noise,
          weight_diff = w_diff,
          attention_diff = mean(trial_att_diffs)
        )
      )

      # If weight difference is 0.5, extract fixation development data
      if (abs(w_diff - 0.5) < 0.001) {
        # Extract fixation development for each trial
        trial_fix_data <- map_dfr(sim$raw, function(trial) {
          fix_seq <- trial$fix_sequence
          max_fix <- min(length(fix_seq), max_fixations)

          # For each fixation position, calculate proportion to attribute 1
          fix_props <- map_dfr(1:max_fix, function(fix_num) {
            # Only use data up to current fixation
            curr_fixes <- fix_seq[1:fix_num]

            # Calculate proportion to attribute 1
            att_indices <- ceiling(curr_fixes / 2)
            att1_prop <- sum(att_indices == 1) / length(att_indices)

            tibble(
              trial = trial$trial,
              fixation_num = fix_num,
              att1_prop = att1_prop
            )
          })

          fix_props
        })

        # Average across trials for each fixation position
        fix_dev_data <- trial_fix_data %>%
          group_by(fixation_num) %>%
          summarize(att1_prop = mean(att1_prop)) %>%
          mutate(noise_level = noise)

        # Add to fixation development results
        fixation_development <- bind_rows(
          fixation_development,
          fix_dev_data
        )
      }
    }
  }

  # Return both datasets
  list(
    weight_att = weight_att_results,
    fix_dev = fixation_development
  )
}

# Create color palette (green gradient as in MATLAB)
create_color_palette <- function(noise_levels) {
  start_color <- c(194, 218, 184) / 255
  end_color <- c(1, 50, 32) / 255

  # Generate color gradient
  colors <- tibble(
    noise_level = noise_levels,
    r = seq(start_color[1], end_color[1], length.out = length(noise_levels)),
    g = seq(start_color[2], end_color[2], length.out = length(noise_levels)),
    b = seq(start_color[3], end_color[3], length.out = length(noise_levels))
  )

  # Convert to hex colors
  colors <- colors %>%
    mutate(hex = rgb(r, g, b))

  # Return as named vector
  setNames(colors$hex, colors$noise_level)
}

# Plot the results
plot_weight_attention_results <- function(results, colors) {
  # Panel A: Fixation development over time (when weight diff = 0.5)
  p1 <- results$fix_dev %>%
    filter(fixation_num <= 20) %>%  # Limit to first 12 fixations
    ggplot(aes(x = fixation_num, y = att1_prop, color = factor(noise_level))) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Sampling Noise") +
    labs(
      x = "Fixation Number",
      y = "p(Fix)_Most Important"
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank()
    ) +
    ylim(0, 1)

  # Panel B: Weight difference vs attention difference
  p2 <- results$weight_att %>%
    ggplot(aes(x = weight_diff, y = attention_diff, color = factor(noise_level))) +
    geom_line() +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, color = "gray70") +
    geom_vline(xintercept = 0.50, color = "gray70") +
    scale_color_manual(values = colors, name = "Sampling Noise") +
    labs(
      x = "Weight_Att1 - Weight_Att2",
      y = "p(Fix)_Att1 - p(Fix)_Att2"
    ) +
    theme_classic() +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    ) +
    ylim(-0.01, 1.01) +
    xlim(-0.01, 1.01)

  # Combine plots
  combined_plot <- p1 + p2 +
    plot_layout(widths = c(1, 1)) +
    plot_annotation(
      title = "Attribute Weights and Attention with Varying Sampling Noise",
      subtitle = "MASC Model Simulation",
      theme = theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    ) &
    theme(legend.position = "bottom")

  combined_plot
}
```

```{r}
# Run the simulation
set.seed(2025)
noise_levels <- seq(0.5, 3, length.out = 3)
weight_diffs <- seq(0, 1, by = 0.25)

# Create color palette
color_palette <- create_color_palette(noise_levels)

# Run simulation (this may take some time)
results <- simulate_weight_attention_relationship(
  n_trials = 30,
  weight_diffs = weight_diffs,
  noise_levels = noise_levels,
  alpha = 10,  # High search sensitivity as in MATLAB
  delta = 0.05 # Same as MATLAB
)
```

```{r, fig.width=12, fig.height=8, out.width="100%"}
# Plot results
fig9_plot <- plot_weight_attention_results(results, color_palette)

# Display plot
print(fig9_plot)
```
