---
title: "Simulating Time Pressure with Varying Levels of Search Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Time Pressure with Varying Levels of Search Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(dplyr)
library(tibble)
library(purrr)
library(patchwork)
library(masc)
```

(A) Payne Index as a function of time pressure (i.e., threshold parameter) and search
sensitivity; at high levels of search sensitivity, a negative relationship between the Payne Index
and time pressure emerges. Variability in the proportion of fixations spent on each attribute (B)
and option (C) increases with time pressure. (D) The proportion of attention directed to the
most important attribute increases with time pressure.

```{r}
# masc_time_pressure_final.R - Creates Figure 10 reproduction from
# Gluth, S., Deakin, J., & Rieskamp, J. (2026). A theory of multiattribute search and choice. Psychological Review.

# Custom function to generate non-dominated attributes
generate_valid_attributes <- function(n_options, n_attributes, lambda = 1) {
  # Keep generating until we get a valid matrix with no dominating options
  while(TRUE) {
    # Generate random values
    x <- matrix(rnorm(n_options * n_attributes), nrow = n_options)

    # Check if any option dominates (is better on all attributes)
    dominates <- logical(n_options)
    for(i in 1:n_options) {
      better_count <- 0
      for(j in 1:n_attributes) {
        if(all(x[i,j] >= x[-i,j])) better_count <- better_count + 1
      }
      dominates[i] <- (better_count == n_attributes)
    }

    # If no option dominates, return this matrix
    if(!any(dominates)) return(x)
  }
}

# Function to simulate MASC behavior under time pressure
simulate_time_pressure_effects <- function(
    n_trials = 50,                     # Number of trials per condition
    n_subjects = 25,                   # Number of simulated participants
    threshold_levels = seq(0.001, 0.2, length.out = 10), # Threshold parameters
    sensitivity_levels = seq(0, 3, length.out = 5),      # Search sensitivity levels
    n_options = 2,                     # Number of options
    n_attributes = 3,                  # Number of attributes
    max_fixations = 100                # Maximum number of fixations
) {
  # Pre-allocate results
  results <- tibble()

  # Create common attribute values and weights for all conditions
  set.seed(2025)

  # Generate attribute weights for each subject
  weights_list <- lapply(1:n_subjects, function(s) {
    # Sample from beta distribution as in the MASC paper
    w <- rbeta(n_attributes, 0.75, 0.75)
    # Normalize to sum to 1
    w / sum(w)
  })

  # Generate attribute values for all trials and subjects
  attr_values <- lapply(1:n_subjects, function(s) {
    # Create dataset for this subject
    lapply(1:n_trials, function(t) {
      # Generate valid attribute values (no option dominates)
      valid_matrix <- generate_valid_attributes(n_options, n_attributes)
      return(valid_matrix)
    })
  })

  # Loop through sensitivity levels
  for(sensitivity_idx in seq_along(sensitivity_levels)) {
    search_sensitivity <- sensitivity_levels[sensitivity_idx]
    cat(sprintf("Processing search sensitivity %.2f (%d/%d)\n",
                search_sensitivity, sensitivity_idx, length(sensitivity_levels)))

    # Loop through threshold levels (time pressure)
    for(thresh_idx in seq_along(threshold_levels)) {
      threshold <- threshold_levels[thresh_idx]
      cat(sprintf("  Processing threshold %.3f (%d/%d)\n",
                  threshold, thresh_idx, length(threshold_levels)))

      # Subject-level metrics
      subject_metrics <- tibble(
        subject = 1:n_subjects,
        payne_index = numeric(n_subjects),
        attr_variance = numeric(n_subjects),
        opt_variance = numeric(n_subjects),
        most_important_prop = numeric(n_subjects),
        choice_consistency = numeric(n_subjects),
        avg_fixations = numeric(n_subjects),
        reward_rate = numeric(n_subjects)
      )

      # Process each subject
      for(s in 1:n_subjects) {
        # Set up parameters for this subject
        weights <- weights_list[[s]]
        most_important <- which.max(weights)

        # Convert attribute values to dataframe format for rMASC
        subject_data <- do.call(rbind, lapply(1:n_trials, function(t) {
          trial_matrix <- attr_values[[s]][[t]]
          as.data.frame(matrix(trial_matrix, nrow = 1))
        }))

        # Rename columns appropriately for rMASC
        colnames(subject_data) <- c(
          paste0("opt", rep(1:n_options, each = n_attributes),
                 "_att", rep(1:n_attributes, n_options))
        )

        # Run rMASC for this subject with current parameters
        result <- rMASC(
          data = subject_data,
          w = weights,
          sigma = 1,  # Fixed sampling noise
          alpha = search_sensitivity,
          delta = 0.01,  # Fixed threshold increase
          theta = threshold,
          max_steps = max_fixations
        )

        # Extract metrics

        # 1. Use the correctness directly from rMASC
        subject_metrics$choice_consistency[s] <- mean(result$results$correct)

        # 2. Average number of fixations
        subject_metrics$avg_fixations[s] <- mean(result$results$rt)

        # 3. Calculate reward rate
        subject_metrics$reward_rate[s] <- subject_metrics$choice_consistency[s] /
          subject_metrics$avg_fixations[s]

        # Extract fixation metrics
        prop_fix_attr <- matrix(0, nrow = n_trials, ncol = n_attributes)
        prop_fix_opt <- matrix(0, nrow = n_trials, ncol = n_options)
        payne_indices <- numeric(n_trials)

        for(t in 1:n_trials) {
          fix_seq <- result$raw[[t]]$fix_sequence

          # Calculate fixation proportions
          for(a in 1:n_attributes) {
            attr_indices <- which(ceiling(fix_seq / n_options) == a)
            prop_fix_attr[t, a] <- length(attr_indices) / length(fix_seq)
          }

          for(o in 1:n_options) {
            opt_indices <- which(((fix_seq - 1) %% n_options) + 1 == o)
            prop_fix_opt[t, o] <- length(opt_indices) / length(fix_seq)
          }

          # Calculate Payne Index for this trial
          within_option <- 0
          within_attribute <- 0

          if(length(fix_seq) > 1) {
            for(i in 1:(length(fix_seq) - 1)) {
              curr_oap <- fix_seq[i]
              next_oap <- fix_seq[i+1]

              curr_opt <- ((curr_oap - 1) %% n_options) + 1
              next_opt <- ((next_oap - 1) %% n_options) + 1

              curr_attr <- ceiling(curr_oap / n_options)
              next_attr <- ceiling(next_oap / n_options)

              if(curr_opt == next_opt) {
                within_option <- within_option + 1
              } else if(curr_attr == next_attr) {
                within_attribute <- within_attribute + 1
              }
            }

            total_transitions <- within_option + within_attribute
            if(total_transitions > 0) {
              payne_indices[t] <- (within_option - within_attribute) / total_transitions
            }
          }
        }

        # Aggregate metrics across trials
        subject_metrics$payne_index[s] <- mean(payne_indices, na.rm = TRUE)
        subject_metrics$attr_variance[s] <- mean(apply(prop_fix_attr, 1, var), na.rm = TRUE)
        subject_metrics$opt_variance[s] <- mean(apply(prop_fix_opt, 1, var), na.rm = TRUE)
        subject_metrics$most_important_prop[s] <- mean(prop_fix_attr[, most_important], na.rm = TRUE)
      }

      # Add aggregated results to main dataframe
      results <- bind_rows(results, tibble(
        search_sensitivity = search_sensitivity,
        threshold = threshold,
        payne_index = mean(subject_metrics$payne_index, na.rm = TRUE),
        attr_variance = mean(subject_metrics$attr_variance, na.rm = TRUE),
        opt_variance = mean(subject_metrics$opt_variance, na.rm = TRUE),
        most_important_prop = mean(subject_metrics$most_important_prop, na.rm = TRUE),
        choice_consistency = mean(subject_metrics$choice_consistency, na.rm = TRUE),
        fixations = mean(subject_metrics$avg_fixations, na.rm = TRUE),
        reward_rate = mean(subject_metrics$reward_rate, na.rm = TRUE)
      ))
    }
  }

  return(results)
}

# Plot results function
plot_time_pressure_results <- function(results) {
  # Create the green gradient color palette
  create_color_palette <- function(n) {
    start_color <- c(194, 218, 184) / 255
    end_color <- c(1, 50, 32) / 255

    # Generate color gradient
    colors <- tibble(
      level = 1:n,
      r = seq(start_color[1], end_color[1], length.out = n),
      g = seq(start_color[2], end_color[2], length.out = n),
      b = seq(start_color[3], end_color[3], length.out = n)
    )

    # Convert to hex colors
    colors <- colors %>%
      mutate(hex = rgb(r, g, b))

    # Return as named vector
    colors$hex
  }

  # Get number of sensitivity levels
  n_sens <- length(unique(results$search_sensitivity))
  colors <- create_color_palette(n_sens)

  # Prepare for plotting by converting sensitivity to factor
  plot_data <- results %>%
    mutate(search_sensitivity = factor(search_sensitivity))

  # Create common theme
  theme_tp <- theme_classic() +
    theme(
      plot.title = element_text(size = 10, face = "bold", hjust = 0),
      axis.title = element_text(size = 9),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      panel.grid.minor = element_blank()
    )

  # Panel A: Payne Index
  p1 <- plot_data %>%
    ggplot(aes(x = threshold, y = payne_index, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "A",
      x = "Threshold θ",
      y = "Payne Index"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Panel B: Attribute Variance
  p2 <- plot_data %>%
    ggplot(aes(x = threshold, y = attr_variance, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "B",
      x = "Threshold θ",
      y = "Attribute Variance"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Panel C: Option Variance
  p3 <- plot_data %>%
    ggplot(aes(x = threshold, y = opt_variance, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "C",
      x = "Threshold θ",
      y = "Option Variance"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Panel D: Most Important Attribute
  p4 <- plot_data %>%
    ggplot(aes(x = threshold, y = most_important_prop, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "D",
      x = "Threshold θ",
      y = "p(Fix = Most Important)"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Additional Panels (optional)
  # Panel E: Choice Consistency
  p5 <- plot_data %>%
    ggplot(aes(x = threshold, y = choice_consistency, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "E",
      x = "Threshold θ",
      y = "Choice Consistency"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Panel F: Number of Fixations
  p6 <- plot_data %>%
    ggplot(aes(x = threshold, y = fixations, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "F",
      x = "Threshold θ",
      y = "Number of Fixations"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Panel G: Reward Rate
  p7 <- plot_data %>%
    ggplot(aes(x = threshold, y = reward_rate, color = search_sensitivity)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = colors, name = "Search Sensitivity") +
    labs(
      title = "G",
      x = "Threshold θ",
      y = "Reward Rate"
    ) +
    theme_tp +
    xlim(-0.05, 0.25)

  # Combine main four panels
  main_plot <- (p1 + p2) / (p3 + p4) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")

  # Combine all seven panels
  full_plot <- (p1 + p2 + p3 + p4) / (p5 + p6 + p7 + plot_spacer()) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")

  list(
    main_plot = main_plot +
      plot_annotation(
        title = "Simulating Time Pressure with Varying Levels of Search Sensitivity",
        theme = theme(plot.title = element_text(size = 14, face = "bold"))
      ),
    full_plot = full_plot +
      plot_annotation(
        title = "Simulating Time Pressure with Varying Levels of Search Sensitivity",
        theme = theme(plot.title = element_text(size = 14, face = "bold"))
      )
  )
}
```

```{r}
# Run simulation (note: this will take some time to run)
set.seed(2025)
results <- simulate_time_pressure_effects(
  n_trials = 200,          # Number of trials
  n_subjects = 100,        # Number of subjects
  threshold_levels = seq(0.001, 0.2, length.out = 10),
  sensitivity_levels = seq(0, 3, length.out = 5)
)
```

```{r, fig.width=12, fig.height=8, out.width="100%"}
# Plot results
plots <- plot_time_pressure_results(results)

# Display main plot (4 panels as in Figure 10)
print(plots$main_plot)

# Display full plot (all 7 panels)
print(plots$full_plot)
```

